Research Experience

### 2010-03-30 doc/thesis/thesis.tex
need to explain the weighting function.

\subsubsection{Resampling}
As a consequence 
of the wide prior distribution (required for a proper discretization of a continuous
distribution), there will be many particles with insignificant weights. While this does help
describe the tails of the distribution very well, it means that only a small portion of the
computation will be spent describing the most probable region. Ideally every particle would 
equally decrease the entropy of the distribution, thus the lower the variance of the weights,
the more efficiently the discrete distribution is in describing the continuous distribution. 
A common measure of "Particle Degeneracy" is the effective number of particles, described
in (Bergman "Navigation and Tracking Applications", 1999, J S Liu and R Chen "Sequential 
Monte Carlo Methods for Dynamical Systems", 1998), which is based on the "true weight"
of each particle. Of course the true weight is unknown, so a heuristic approximating 
$N_{eff}$ is used:
\begin{equation}
N_{eff} \approx \frac{N_p}{\sum_{i=1}^{N_p} w_i^2}
\end{equation}
Any quick run of a particle filter will reveal that unless the prior is particularly accurate,
$N_{eff}$ drops precipitously.  To alleviate this problem
a common technique known as resampling must be applied. The idea of re-sampling is to 
draw from the approximate posterior, thus generating a replica of the posterior with 
a support more suited to the posterior. Thus, if weights are all set to $1/N_p$, and 
$N_p$ points are drawn from the posterior,
\begin{equation}
\hat{x}_j \sim \left(\sum_{i=1}^{N_p} w_i(t)\delta(x - x_i(t))\right), j \in \{1, ..., N_p\}
\end{equation}
then $\hat{x} \sim x$. Unfortunately this isn't necessarily the truth: since the support is
still limited to the original particles, the number of unique particles may only go down.
This effect, often dubbed "particle impoverishment" can result in excessive quantization
errors in the final distribution.

%<more here>

@@

%<stopped here>
This has the added benefit that it is not necessary to
re-draw points from a $q$ function at every time step, which is faster. 
this the definition of the weight, so when weights are updated it is necessary
to divide by the proposal distributionto converge
to the true distribution it is necessary to divide by the proposal distribution.

Thus $\hat{x}(0)$ is an empirical representation of the prior, and $\hat{x}_i(0)$
is the $i^{th}$ particle, a single possible representation of the system at time 0.
Initially, since each $\hat{x}_i$ has been drawn from the prior, and no measurements
have been made, the particles are all equally weighted. When a measurement becomes
available, that information is then incorporated into the posterior by re-weighting
the particles according to how well they estimated this new measurement.

The estimated posterior of $\hat{x}(t)$, for discrete time, $t$ (before incorporating
any measurement that might be available at $t$) is then:
\begin{equation}
\Pr\{\hat{x}(t) | y[0:t-1]\} =  \int \Pr\{\hat{x}(t), \hat{x}(t-1) | y[0:t-1]\} d\hat{x}(t-1)
\end{equation}

and by bayes theorem $\Pr\{\hat{x}(t), \hat{x}(t-1) | y[0:t-1]\} = 
\Pr\{\hat{x}(t) | \hat{x}(t-1), y[0:t-1]\}\Pr\{\hat{x}(t-1) | y[0:t-1]\}$,
therefore 
\begin{equation}
\Pr\{\hat{x}(t) | y[0:t-1]\} =  \int \Pr\{\hat{x}(t) | \hat{x}(t-1) ,y[0:t-1]\}\Pr\{\hat{x}(t-1) | y[0:t-1]\}  d\hat{x}(t-1)
\end{equation}

Since in this work we will only deal with markov processes, we assume that 
$\Pr\{\hat{x}(t) | \hat{x}(t-1), y[0:t-1]\} = \Pr\{\hat{x}(t) | \hat{x}(t-1)\}$

thus the posterior becomes:
\begin{equation}
\Pr\{\hat{x}(t) | y[0:t-1]\} =  \int \Pr\{\hat{x}(t) | \hat{x}(t-1)\}\Pr\{\hat{x}(t-1) | y[0:t-1]\}  d\hat{x}(t-1)
\end{equation}

Therefore in between measurements, there is no need to know previous measurements in
order to make predictions about the internal state. This is important because it means makes
particle filters significantly more memory efficient, and although it isn't useful in this
case, it means a particle filter could theoretically be run indefinitely without the
need to store old data.

As stated previously, new measurements may be incorporated by reweighting particles.
To find the updated distribution, $\Pr\{\hat{x}(t) | y[0:t]\}$, bayes theorem is
again applied:
\begin{equation}
\Pr\{\hat{x}(t) | y[0:t]\} = \Pr\{\hat{x}(t) | y[0:t-1], y(t)\} = \frac{\Pr\{y(t) | \hat{x}(t), y[0:t-1]\}\Pr\{\hat{x}(t) | y[0:t-1]\}}
{\Pr\{y(t) | y[0:t-1]\}}
\end{equation}

\begin{equation}
\Pr\{\hat{x}_i(t) | y[0:t-1]\} = \int \Pr\{\hat{x}_i(t) | \hat{x}_i(t-1), y[0:t-1]\}\Pr[\hat{x}_i(t-1)
\end{equation}
Since $\hat{x}(t) = \hat{x}(t-1) + \int_{t-1}^t f(x(t), u(t))dt + \int_{t-1}^t\nu_d dt$, 
\begin{equation}
\Pr\{\hat{x}_i(t) | \hat{x}_i(t-1)\} = \Pr\{\int{t-1}^t \nu_d\}
\end{equation}

of course $\int_{t-1}^t f(x(t), u(t))dt$ may need to be evaluated numerically. Therefore,
to estimate the posterior probability at $t$, it is just necessary to propagate
each particle forward in time according to a state space function. To form 
the posterior probability around the true $x(t)$, each particle is weighted
based on

This
then results in the expected value of $\hat{y}(t)$:
\begin{equation}
E[\hat{y}_i(t) | \hat{x}_i(t-1)] = E[\hat{x}_i(t)]
\end{equation}

\begin{equation}
\dot{y} = -ky
\end{equation}
$\theta_i$ would have an estimate for $k$ \emph{and} $y$. Of course in many cases
the model is a potentially non-linear differential equation with no closed
form solution. Initially every $\theta_i$ will have the same weight, because
there isn't yet anything to base a weighting on. Note that this doesn't mean
that the particles have been evenly distributed, so the distribution is not 
necessarily "flat" just because all the particles are equaly weighted. When a
new measurement is available, the particles are then reweighted based on how
well they predicted the new measurement.

\subsection{My Specific Particle Filte}
Although particle filters are well defined, significant design decisions
are application dependent. The most important choice is the weighting function,
$w(y_t - y(\theta_t))$. Obviously the function should weight $0$ maximally,
the drop off rate is extremely important. A very thin gaussian probability 
distribution function has nice properties, but thin tails. As a result, large
outliers in the measurement vector could easily force all the particles to
have near 0 weights, thus forcing the particle filter to converge improperly. 
On the other end of the spectrum, a cauchy PDF may not
weight particles in the middle enough, preventing the particles to never 
converge. Of course, the scale factor, or variance, also plays a signifcant
role in the convergeance rate. The importance of the weighting function
cannot be overstated, as this is the primary factor in deciding the rate at
which the particle filter converges. As part of the simulations, I compare
a gaussian based weighting function with an exponential weightin function.
In both cases, I scaled the width of the weighting functions automatically based
on the signal variance.

>>>>

\subsubsection{Resampling}
As a consequence 
of the wide prior distribution (required for a proper discretization of a continuous
distribution), there will be many particles with insignificant weights. While this does help
describe the tails of the distribution very well, it means that only a small portion of the
computation will be spent describing the most probable region. Ideally every particle would 
equally decrease the entropy of the distribution, thus the lower the variance of the weights,
the more efficiently the discrete distribution is in describing the continuous distribution. 
A common measure of "Particle Degeneracy" is the effective number of particles, described
in (Bergman "Navigation and Tracking Applications", 1999, J S Liu and R Chen "Sequential 
Monte Carlo Methods for Dynamical Systems", 1998), which is based on the "true weight"
of each particle. Of course the true weight is unknown, so a heuristic approximating 
$N_{eff}$ is used:
\begin{equation}
\label{neff}
N_{eff} \approx \frac{N_p}{\sum_{i=1}^{N_p} w_i^2}
\end{equation}
Any quick run of a particle filter will reveal that unless the prior is particularly accurate,
$N_{eff}$ drops precipitously.  To alleviate this problem
a common technique known as resampling must be applied. The idea of re-sampling is to 
draw from the approximate posterior, thus generating a replica of the posterior with 
a support more suited to the posterior. Thus, if weights are all set to $1/N_p$, and 
$N_p$ points are drawn from the posterior,
\begin{equation}
\hat{x}_j \sim \left(\sum_{i=1}^{N_p} w_i(t)\delta(x - x_i(t))\right), j \in \{1, ..., N_p\}
\end{equation}
then $\hat{x} \sim x$. Unfortunately this isn't necessarily the truth: since the support is
still limited to the original particles, the number of unique particles may only go down.
This effect, often dubbed "particle impoverishment" can result in excessive quantization
errors in the final distribution. However, by instead sampling from a smoothed version of
the discrete distribution, sample impoverishment is prevented. The question then is how to 
decide on the smoothing kernel. 
Often times the easiest way to sample from the continous distribution is to break the 
re-sampling down into two steps. First a member of the discrete distribution is randomly
selected based on the weights, and then based on the smoothing a nearby state variable 
is selected. The process of the selection will be defined as:
\begin{equation}
x'_i = x_i + h\sigma \epsilon
\end{equation}
Where $h$ is the bandwidth, $\sigma$ is the standard deviation such that $\sigma \sigma^T = cov(x)$
and $\epsilon$ is drawn from the chosen kernel.
It has been proven that when all the elements of the mixture
have the same weight, as is the case after basic re-sampling, the kernel that minimizes the 
MSE between the estimated and true posterior is the Epanechnikov Kernel (cite Improving Regularised
Particle Filters, C Musso, N Oudjane and F LeGrand). 
\begin{equation}
K = \left\{
\begin{array}{lr}
\frac{n_x+2}{2c_{n_x}}(1-\|x\|^2) & if\ \|x\| < 1\\
0 & otherwise
\end{array}\right.
\end{equation}
%<more here>

If the noise is assumed to be Gaussian then it is possible to further optimize. Obviously,
its very possible that distribution will deviate from that assumption, however in this
case assuming a Gaussian still allows a great deal of flexibility, since the distribution
is a mixture. Thus we let $h$ be defines as:
\begin{eqnarray}
h = [N_s8c^{-1}_{n_x}(n_x + 4)(2\sqrt{\pi})^{n_x}]^{\frac{1}{n_x +4}}
\end{eqnarray}
and although it is very possible the underlying noise is non-gaussian, the Gaussian
may still be a good choice. Its worth noting that as $n_x$, the number of dimensions in
$x$ go to infinity, the standard deviation based approximation becomes inefective according
 to (cite a Tutorial on Particle Filters for on-line non-linear non-gaussian bayesian
 tracking, sanjeev arulampalam, simon maskell, neil gordon...). Because our regularized 
 resampling method is based on particle variance, resampling very often will slow down 
 convergeance, because the resampling itself can lead to increased variance. Thus, for
 the most part the resulting algorith is the basic particle filter, except for when the
 $N_{eff}$ drops very low (say less than 50), at which point regularized resampling is 
 performed. The resulting algorthm is as follows:

 \begin{algorithmic}
\STATE Initialize $N_p$ Particles: 
        $\{x_i(0),w_i(0) : x_i(0) \sim \alpha(x), w_i(0) = \frac{1}{N_p}, i \in \{1, 2, ... , N_p\} \}$
\STATE $T$ = \{Set of Measurement Times\}
\FOR{$t$ in $T$}
    \FOR{$i$ in $N_p$}
        \STATE $x_i(t) = x_i(t-1) + \int_{t-1}^t f(x(\tau), u(\tau)) d\tau $
        \STATE $w_i(t) = w_i(t-1)\hat{Pr}(y(t) | x(t))$
    \ENDFOR

    \STATE Calculate $N_{eff}$ with \eqref{neff}
    \IF{$N_{eff} < N_R$ (recommend $N_R = min(50, .1N_p)$ )}
        \STATE Calculate empirical $\sigma$ 
        \STATE Redraw particles using basic resampling
        \FOR{$i$ in $N_p$}
            \STATE Draw $\epsilon \sim K$
            \STATE $x_i = x_i + h \sigma \epsilon$
        \ENDFOR
    \ENDIF
\ENDFOR

\STATE At $t + \Delta t$, $t \in T$, $\hat{Pr}(x(t+\Delta t)) \approx 
\sum_{i=1}^{N_p} w_i(t)\delta\left(x - (x_i(t) + \int_t^{t+\Delta t} f(x(\tau), u(\tau)) d\tau) \right)$

 \end{algorithmic}

@@

\subsubsection{Choosing $\hat{Pr}(y(T) | x(T))$}
Analysis of Gaussian versus exponential

Discussion of optimality (or lack thereof)

Discussion of using a gaussian mixture pdf to describe a non-gaussian distribution

%<stopped here>
\subsection{My Specific Particle Filte}
Although particle filters are well defined, significant design decisions
are application dependent. The most important choice is the weighting function,
$w(y_t - y(\theta_t))$. Obviously the function should weight $0$ maximally,
the drop off rate is extremely important. A very thin gaussian probability 
distribution function has nice properties, but thin tails. As a result, large
outliers in the measurement vector could easily force all the particles to
have near 0 weights, thus forcing the particle filter to converge improperly. 
On the other end of the spectrum, a cauchy PDF may not
weight particles in the middle enough, preventing the particles to never 
converge. Of course, the scale factor, or variance, also plays a signifcant
role in the convergeance rate. The importance of the weighting function
cannot be overstated, as this is the primary factor in deciding the rate at
which the particle filter converges. As part of the simulations, I compare
a gaussian based weighting function with an exponential weightin function.
In both cases, I scaled the width of the weighting functions automatically based
on the signal variance.
